Exploring probability distributions for bivariate temporal granularities

### 2019-11-25 main.Rmd
spell check rough draft

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in [@Wickham2009-pk]. 
bibliography: [bibliography.bib]
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

@@

Temporal data are available at various resolution depending on the context. Social and economic data like GDP are often collected and reported at coarse temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be recorded every second. As the frequency of data increases, the number of questions about periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities[@aigner2011visualization].
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore all possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

@@

Calendar-based graphics[@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on arrangement (linear vs. cyclic) or hierarchical order of calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in [@Bettini1998-ed]. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular or aperiodic depending on if mappings between granularities are regular or irregular. Examples of circular can be hour of the day, day of the week and that of aperiodic granularities can be day of the month or public holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in household across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates of usage [@2012-la] across households, total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data can be benefited from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. However, the motivation came through the smart meter example, this is a problem which relates to any data that needs to be analysed for different periodicities.

@@

# Definitions of time granularities{#sec:timegrandef}

@@

Providing a formalism to some of these abstractions is important to model a time series across differently grained temporal domains.

@@

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_(R-1)}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_(R-1)}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H (j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$.

@@

\begin{equation} \label{eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi
\end{equation}
where,  
z denotes the index set.  
BG denotes the index of the tsibble (bottom granularity).  
G be a linear granularity, periodic with respect to BG.    
k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.

@@

All circular granularities can be expressed in terms of the bottom linear granularity with equation \ref{eq2}.

@@

An **Aperiodic granularity** can not be defined using modular arithmetic. The modulus for these type of calendar categorizations are not constant due to irregular mapping with bottom granularities. For example, table \ref{tab:aperiodic} shows some example of aperiodic granularities ($A_i$) and plausible choices of period lengths($k_i$). The coarse granularity is assumed as the period within which the finer granularity will repeat its behavior aperiodically.

@@

**Discussion**: Example showing aperiodic granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:aperiodic-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $AG = Alter^{5}_{(1,-1)}(BG, Group_3(BG))$ and $AH = Alter^{12}_{(1,-2)}(BG, Group_7(BG))$. This implies that AG is made up of shrinking every 1st granule of Group_3(BG) by 1 and AH is made up of shrinking every 1st granule of Group_3(BG) by 2. Number of granules of AG and AH in each period of BG is 2, but the number of granules of BG in each of those granules are different. A(BG, AG) and A(BG, AH) are repetitive categorization of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of AG or AH. For AG, R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For AH, R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \ref{eq:eq4} and  \ref{eq:eq5}.

@@

The linear granularities G, H, AG and AH considered in the definitions of circular and aperiodic granularities are all periodic with respect to the bottom granularity. The only difference is G, H have regular mapping, whereas, G1 and G2 have irregular mapping with respect to BG. It can be noted here that if a linear granularity AG is quasi-periodic with respect to BG, then \ref{eq:eq3} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$ (\autoref{def:quasi})

@@

but the grouping is always periodic.

@@

In \ref{sec:singlefrommultiple} and \ref{sec:multiplefromsingle}, we assume that all single order granularities are circular.What happens when we have a mix of circular and aperiodic granularities.

@@

The single order-up granularites resulting from the first two cases are periodic and has been handled in \ref{sec:singlefrommultiple} and \ref{sec:multiplefromsingle}. The calendar categorization resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. Let us consider the computation of $A_{hour, month}$ given single order-up granularities $A_{hour, day} = 12$, $A_{day, month} = 20$ and $C_{month, year} = 9$.

@@

In Figure \ref{fig:day-fortnight}, the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy behavior of these two customers. it can be seen that for most days of the fortnight, the second household has much less consumption than the first one. However, there is additional information which we can derive looking at the distribution. If we consider letter value F as a regular behavior and letter values beyond F as not-so-regular behavior, we can conclude that the regular behavior of the first household is more stable than the second household. However, the distribution of tail of the first household is more variable, observed through distinct letter values, implying that their not-so-regular behavior is quite extreme. This shows, how looking at the distribution of the dependent variable can throw more light on the energy behavior of the customers, which are lost using aggregate or summary statistics.

@@

In \ref{fig:bothcust} we visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across wknd_wday (facet) and hour-day (x-axis) for the same two households. For the second household, outliers are less prominent implying their regular behavior is more stable. For the first household, energy behavior is not significantly different between weekdays and weekends. For the second household, median energy consumption for the early morning hours is extremely high for weekends compared to weekdays.

>>>>

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2[@Wickham2009-pk]. 
bibliography: [bibliography.bib]
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

@@

Temporal data are available at various resolutions depending on the context. Social and economic data like GDP is often collected and reported at coarse temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities[@aigner2011visualization].
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore all possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

@@

Calendar-based graphics[@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs. cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in [@Bettini1998-ed]. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular or aperiodic depending on if mappings between granularities are regular or irregular. Examples of circular can be hour of the day, day of the week and that of aperiodic granularities can be day of the month or public holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates of usage across households (for example, [@2012-la]), total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data can be benefited from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. However, the motivation came through the smart meter example, this is a problem which relates to any data that needs to be analyzed for different periodicities.

@@

# Time granularities{#sec:timegrandef}

@@

In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.

@@

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_(R-1)}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_{R-1}}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H (j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$.

@@

\begin{equation} \label{eq:eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi
\end{equation}
where,  
z denotes the index set.  
BG denotes the index of the tsibble (bottom granularity).  
G be a linear granularity, periodic with respect to BG.    
k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.

@@

All circular granularities can be expressed in terms of the bottom linear granularity with equation \autoref{eq:eq2}.

@@

An **Aperiodic granularity** can not be defined using modular arithmetic. The modulus for these type of calendar categorizations are not constant due to irregular mapping with bottom granularities. For example,\autoref{tab:aperiodic} shows some example of aperiodic granularities ($A_i$) and plausible choices of period lengths($k_i$). The coarse granularity is assumed as the period within which the finer granularity will repeat its behavior aperiodically.

@@

**Discussion**: Example showing aperiodic granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:aperiodic-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $AG = Alter^{5}_{(1,-1)}(BG, Group_3(BG))$ and $AH = Alter^{12}_{(1,-2)}(BG, Group_7(BG))$. This implies that AG is made up of shrinking every 1st granule of Group_3(BG) by 1 and AH is made up of shrinking every 1st granule of Group_3(BG) by 2. Number of granules of AG and AH in each period of BG is 2, but the number of granules of BG in each of those granules are different. A(BG, AG) and A(BG, AH) are repetitive categorization of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of AG or AH. For AG, R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For AH, R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \autoref{eq:eq4} and  \autoref{eq:eq5}.

@@

The linear granularities G, H, AG and AH considered in the definitions of circular and aperiodic granularities are all periodic with respect to the bottom granularity. The only difference is G, H have regular mapping, whereas, G1 and G2 have irregular mapping with respect to BG. It can be noted here that if a linear granularity AG is quasi-periodic with respect to BG, then \autoref{eq:eq3} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$ (\autoref{def:quasi})

@@

@@

In \autoref{sec:singlefrommultiple} and \autoref{sec:multiplefromsingle}, we assume that all single order granularities are circular.What happens when we have a mix of circular and aperiodic granularities.

@@

The single order-up granularites resulting from the first two cases are periodic and has been handled in \autoref{sec:singlefrommultiple} and \autoref{sec:multiplefromsingle}. The calendar categorization resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. Let us consider the computation of $A_{hour, month}$ given single order-up granularities $A_{hour, day} = 12$, $A_{day, month} = 20$ and $C_{month, year} = 9$.

@@

In Figure \ref{fig:day-fortnight}, the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy behavior of these two customers. it can be seen that for most days of the fortnight, the second household has much less consumption than the first one. However, there is additional information that we can derive looking at the distribution. If we consider letter value F as a regular behavior and letter values beyond F as not-so-regular behavior, we can conclude that the regular behavior of the first household is more stable than the second household. However, the distribution of tail of the first household is more variable, observed through distinct letter values, implying that their not-so-regular behavior is quite extreme. This shows, how looking at the distribution of the dependent variable can throw more light on the energy behavior of the customers, which are lost using aggregate or summary statistics.

@@

In \autoref{fig:bothcust} we visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across wknd_wday (facet) and hour-day (x-axis) for the same two households. For the second household, outliers are less prominent implying their regular behavior is more stable. For the first household, energy behavior is not significantly different between weekdays and weekends. For the second household, median energy consumption for the early morning hours is extremely high for weekends compared to weekdays.
